### 2020-09-14 manuscript/paper_BRM.Rmd
add general eye movement event classification section to introduction

Most of these classification algorithms use non-parametric methods to differentiate between eye movement events. For instance, the classic "Velocity-Threshold" algorithm classifies samples with a velocity above a fixed threshold as saccades [@Stample1993]. More recently, supervised statistical learning techniques have been successfully used to parse eye movement data. For example, @Bellet2019 trained a convolutional neural network on eye-tracking data from humans and macaques and achieved classification results that were similarly accurate to those of human coders. However, these supervised methods have several disadvantages: First, they require labeled data (usually labeled by humans) as input. Second, they often rely on hyperparameters which are difficult to theoretically justify. Third, their parameters are usually difficult to interpret: Most supervised methods operate in a black-box fashion, which implies that they don't make any explicit assumptions about the data generating process. Oppositely, several algorithms using unsupervised techniques have been developed. They typically don't need to be trained on labeled data and are thus more flexible in classifying data from different individuals, tasks, or eye-trackers.  
Besides differentiating between supervised and unsupervised methods, algorithms can vary in whether they are explicitly modeling the data generating process and are thus able to simulate new data. To our knowledge, these *generative models* have been rarely used to classify eye movement data [cf. @Mihali2017, @Wadehn2019]. However, they come with various advantages and we will summarize the two most important for eye movement classification here [for an overview, see @Epstein2008].  
First, generative models allow to not only predict (classify) but also explain eye-tracking phenomena. For instance, they can inhere parameters that explain why oscillations only occur after but not before saccades. When the goal is to understand eye movements and improve their classification based on this understanding, this aspect is clearly an advantage over non-parametric or supervised methods. Second, generative models can challenge prevailing theoretical assumptions and bring up new research questions. For example, they might suggest that oscillations also occur before saccadic eye movements [as has been mentioned by @Nystrom2010] or that the assumption of discrete eye movements does not hold [as has been discussed in @Andersson2017; maybe others].  

@@

Several studies have tested the performance of HMM algorithms against other classification methods: I-HMM has been deemed as robust against noise, behaviorally accurate, and showing a high sample-to-sample agreement to human coders [@Andersson2017; @Komogortsev2010; @Salvucci2000]. However, the agreement was lower when compared to an algorithm using a Bayesian mixture model [@Kasneci2014; @Tafaj2012]. NSLR-HMM showed even higher agreement to human coding than I-HMM but was outperformed by a recently developed algorithm using convolutional neural networks [@Bellet2019; @Pekkanen2017].  

@@

First, I-HMM relies on setting an appropriate threshold to determine the initial classification, which can distort the results [@Blignaut2009; @Komogortsev2010; @Shic2008]. Second, the current implementation of NSLR-HMM requires human-coded data, which limits its applicability. It also inherits fixed parameters that prevent the algorithm to adapt to the individual- or task-specific signals (contrary to the authors, we argue that this is a disadvantage). Third, BMD limits the classification to microsaccades which are irrelevant in many applications and sometimes even considered as noise [@Duchowski2017]. The opposite problem was observed for BP-AR-HMM: It tends to estimate an unreasonable number of events from the data of which many are considered as noise events. Therefore, the authors suggest using it as an exploratory tool followed by further event classification [@Houpt2018].  

@@

# The Algorithm
## Preprocessing
Algorithms require variables that describe gaze data (hereafter called *eye movement metrics*) to classify them into events. Many eye movement metrics have been proposed and used in previous algorithms [for examples, see @Zemblys2018; @Andersson2017]. However, many of them rely on thresholds or window ranges that have to be set by the user [e.g., the distance between the mean position in a 100 ms window before and after each sample, see @Olsson2007]. This can be problematic because such parameters are often set without theoretical justification and differ substantially between metrics or heavily depend on the eye-tracker's sampling frequency [@Andersson2017]. In gazeHMM, we used velocity, acceleration, and sample-to-sample angle [synonymous to relative or change in angle; @Larsson2013] because they belong to the most basic metrics which do not require parameter settings.  
Theoretically, these three metrics should separate eye movement events clearly. Fixations typically inherit samples with low velocity and acceleration [@Larsson2013]. Due to tremor, we assume that the angle between samples should not follow any direction but a random walk [@Duchowski2017; @Larsson2015]. In contrast, saccade samples usually have a high velocity and acceleration and roughly follow the same direction. PSO samples tend to have moderate velocity and high acceleration since they occur between saccades and low velocity events [@Larsson2013]. They can be specifically distinguished by their change in direction clustered around 180 degrees [@Pekkanen2017]. Importantly, the oscillations depend on the resolution of the gaze recording: Eye-trackers with higher sampling frequency yield more changes in direction and more samples in between those changes. Those samples in between typically follow the same direction. Thus, with high sampling frequencies, PSO samples might also cluster around a sample-to-sample angle of zero with outliers around 180 degrees. Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades they follow the same direction [@Larsson2013; @Leigh2015]. Other algorithms focus exclusively on classifying microsaccades [e.g., @Mihali2017], but as stated earlier, these events were not in the scope of gazeHMM.  
The velocity and acceleration signals are computed from the raw gaze position by using a Savitzky-Golay filter [@Savitzky1964; similar to @Nystrom2010]. The sample-to-sample angle is calculated as: $$\alpha(t) = \arctan\left(\frac{y_{t+1} - y_{t}}{x_{t+1} - x_{t}}\right) -\arctan\left(\frac{y_{t} - y_{t-1}}{x_{t} - x_{t-1}}\right),$$ with $\alpha(t) := \alpha(t) + 2\pi$ for $\alpha(t) < 0$, and is therefore bound between 0 and $2\pi$.  
Most of the missing data in eye movement classification are due to blinks. Therefore, the user can indicate which samples should be labeled as blinks (other missing samples are treated as noise). Often, eye-trackers record a few samples with unreasonably high velocity and acceleration before losing the pupil signal when a blink occurs. Since these samples could distort the classification of saccades in the HMM, gazeHMM removes them heuristically. Before classifying the samples, it sets all samples within 50 ms before and after blink samples as missing. The window of 50 ms was rather motivated empirically than theoretically and can bet set to any value the user considers appropriate.

>>>>

On one hand, many of these classification algorithms use non-parametric methods to differentiate between eye movement events. For instance, the classic "Velocity-Threshold" algorithm classifies samples with a velocity above a fixed threshold as saccades [@Stample1993]. On the other hand, many parametric methods have been developed recently. Some of them require human labeled training data as input and can therefore be termed as supervised. For example, @Bellet2019 trained a convolutional neural network (CNN) on eye-tracking data from humans and macaques and achieved classification results that were similarly accurate to those of human coders. However, these supervised methods have several disadvantages: First, they require labeled data (usually labeled by humans) as input. Second, they often rely on hyperparameters which are difficult to theoretically justify. Third, their parameters are usually difficult to interpret: Most supervised methods operate in a black-box fashion, which implies that they don't make any explicit assumptions about the data generating process. Beyond that, there are several algorithms using unsupervised techniques. They typically don't need to be trained on labeled data and are also more flexible in classifying data from different individuals, tasks, or eye-trackers.  
Besides discriminating between supervised and unsupervised methods, algorithms can vary in whether they are explicitly modeling the data generating process and are thus able to simulate new data. To our knowledge, these *generative models* have been rarely used to classify eye movement data [cf. @Mihali2017, @Wadehn2019]. However, they come with various advantages and we will summarize the two most important ones for eye movement classification here [for an overview, see @Epstein2008].  
First, generative models allow to not only predict (classify) but also explain eye-tracking phenomena. For instance, they can inhere parameters that explain why oscillations only occur after but not before saccades. When the goal is to understand eye movements and improve their classification based on this understanding, this aspect is clearly an advantage over non-parametric or supervised methods. Second, generative models can challenge prevailing theoretical assumptions and bring up new research questions [@Epstein2008]. For example, they might suggest that oscillations also occur before saccadic eye movements [as has been mentioned by @Nystrom2010] or that the assumption of discrete eye movements does not hold [as has been discussed in @Andersson2017].  

@@

Several studies have tested the performance of HMM algorithms against other classification methods: I-HMM has been deemed as robust against noise, behaviorally accurate, and showing a high sample-to-sample agreement to human coders [@Andersson2017; @Komogortsev2010; @Salvucci2000]. However, the agreement was lower when compared to an algorithm using a Bayesian mixture model [@Kasneci2014; @Tafaj2012]. NSLR-HMM showed even higher agreement to human coding than I-HMM but was outperformed by the CNN algorithm by @Bellet2019 [@Pekkanen2017].  

@@

First, I-HMM relies on setting an appropriate threshold to determine the initial classification, which can distort the results [@Blignaut2009; @Komogortsev2010; @Shic2008]. Second, the current implementation of NSLR-HMM requires human-coded data, which narrows its applicability. It also inherits fixed parameters that prevent the algorithm to adapt to the individual- or task-specific signals. Third, BMD limits the classification to microsaccades which are irrelevant in many applications and sometimes even considered as noise [@Duchowski2017]. The opposite problem was observed for BP-AR-HMM: It tends to estimate an unreasonable number of events from the data of which many are considered as noise events. Therefore, the authors suggest using it as an exploratory tool followed by further event classification [@Houpt2018].  

@@

The following section will describe gazeHMM and the underlying HMM in detail. Then, we will present the parameter recovery of the HMM and show how the algorithm performs compared to other eye movement event classification algorithms with regard to the agreement to human coding. Finally, we will discuss these results and propose directions in which gazeHMM and other HMM algorithms could be improved.

# The Algorithm
As illustrated in Figure \@ref(fig:plot-algorithm-workflow), most eye movement event classification algorithms consist of three steps: During *preprocessing*, the raw gaze position is transformed into a metric of position and/or time. Often, a filtering or smoothing procedure is applied to the data before or after the tranformation to separate the gaze signal from noise and artifacts [@Spakov2012]. It follows the *classification*, depending on the method and settings of the algorithm, each sample is labeled as a candidate for one of the predefined events. Lastly, as part of the *postprocessing*[^1], the algorithm decides which candidates to accept, relabel, or merge [@Hessels2017; @Komogortsev2010].

[^1]: @Hessels2017 called step two the *search rule* and step three the *classifcation rule*. For non-parametric methods, this distinction might be accurate. However, for parametric methods, we think calling step two *classification* is more appropriate since the probabilistic classification is done here. Step three usually consist of some heuristic relabeling and correcting for classification errors.

## Preprocessing
Algorithms require variables that describe gaze data (hereafter called *eye movement metrics*) to classify them into events. Many eye movement metrics have been proposed and used in previous algorithms [for examples, see @Zemblys2018; @Andersson2017]. However, many of them rely on thresholds or window ranges that have to be set by the user [e.g., the distance between the mean position in a 100 ms window before and after each sample, see @Olsson2007]. This can be problematic because such parameters are often set without theoretical justification and differ substantially between metrics or heavily depend on the eye-tracker's sampling frequency [@Andersson2017]. In gazeHMM, we used velocity, acceleration, and sample-to-sample angle [synonymous to relative or change in angle; @Larsson2013] because they belong to the most basic metrics which do not require parameter settings.  
Theoretically, these three metrics should separate eye movement events clearly. Fixations typically inherit samples with low velocity and acceleration [@Larsson2013]. Due to tremor, we assume that the angle between samples should not follow any direction but a random walk [@Duchowski2017; @Larsson2015]. In contrast, saccade samples usually have a high velocity and acceleration and roughly follow the same direction. PSO samples tend to have moderate velocity and high acceleration since they occur between saccades and low velocity events [@Larsson2013]. They can be specifically distinguished by their change in direction clustered around 180 degrees [@Pekkanen2017]. Importantly, the oscillations depend on the resolution of the gaze recording: Eye-trackers with higher sampling frequency yield more changes in direction and more samples in between those changes. Those samples in between typically follow the same direction. Thus, with high sampling frequencies, PSO samples might also cluster around a sample-to-sample angle of zero with outliers around 180 degrees. Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades they follow the same direction [@Larsson2013; @Leigh2015]. Other algorithms focus exclusively on classifying microsaccades [e.g., @Mihali2017], but as stated earlier, these events were not in the scope of gazeHMM.  
The velocity and acceleration signals are computed from the raw gaze position by using a Savitzky-Golay filter [@Savitzky1964; similar to @Nystrom2010]. The sample-to-sample angle is calculated as: $$\alpha(t) = \arctan\left(\frac{y_{t+1} - y_{t}}{x_{t+1} - x_{t}}\right) -\arctan\left(\frac{y_{t} - y_{t-1}}{x_{t} - x_{t-1}}\right),$$ with $\alpha(t) := \alpha(t) + 2\pi$ for $\alpha(t) < 0$, and is therefore bound between 0 and $2\pi$.  
Most of the missing data in eye movement classification are due to blinks. Therefore, the user can indicate which samples should be labeled as blinks (other missing samples are treated as noise). Often, eye-trackers record a few samples with unreasonably high velocity and acceleration before losing the pupil signal when a blink occurs. Since these samples could distort the classification of saccades in the HMM, gazeHMM removes them heuristically. Before classifying the samples, it sets all samples within 50 ms before and after blink samples as missing. The window of 50 ms was rather motivated empirically than theoretically and can bet set to any value the user considers appropriate.
